A comparison on R tools for nonlinear least squares modeling

### 2023-05-09 RefactoringNLS/ComparingNLStools.Rmd
finish first pass edit to shorten

```{r exhobbsfn, eval=TRUE}
hobnlfb<-nlfb(start=stu1, resfn=hobbs.res, jacfn=hobbs.jac)
pshort(hobnlfb) # use print(hobnlfb) for more detail
hobnlm<-nls.lm(par=stu1, fn=hobbs.res, jac=hobbs.jac)
pnlm0(hobnlm)  
hobgsln<-gsl_nls(start=stu1, fn=hobbs.res, y=rep(0,12))
pnlm0(hobgsln)
hobgsl<-gsl_nls(start=stu1, fn=hobbs.res, y=rep(0,12), jac=hobbs.jac)
pnlm0(hobgsl) # using analytic Jacobian
```

@@

The design goals of the different tools are also evident in the so-called
"convergence tests" for the iterative solvers. In the manual page for `nls()` 
in R 4.0.0 there was the warning:

@@

with suggested addition of small perturbations to the data. This amounted to 
admitting `nls()` cannot solve well-posed problems unless data is polluted
with errors. We dispute that zero-residual problems are always artificial,
since problems in function approximation and nonlinear equations can be 
approached with nonlinear least squares. Moreover, the matter 
can be easily resolved. The "termination test" for the **program** rather than for 
"convergence" of the underlying **algorithm** is the Relative Offset Convergence
Criterion (see @BatesWatts81). This projects the proposed step in the parameter
vector onto the gradient and estimates how much the sum of squares loss function 
should decrease. This estimate is divided 
by the current size of the loss function to avoid scale issues.
When we have "converged", the estimated relative
decrease is very small. However, with small residuals,
the sum of squares loss function is (almost) zero and we get the possibility of a 
zero-divide failure.

Adding a small quantity to the loss function before dividing avoids trouble. 
In 2021, one of us (J. Nash) proposed that `nls.control()` have an 
additional parameter `scaleOffset` with a default value of zero. Setting it to a small 
number -- 1.0 is a reasonable choice -- allows small-residual problems 
(i.e., near-exact fits) to be dealt with easily. We call this the
**safeguarded relative offset convergence criterion**. The default value 
gives the legacy behaviour.
We are pleased to note that this improvement is now in the R distributed code 
since version 4.1.0; it has been a part of `nlsr` since it was introduced.

More general termination tests can be used. `nlsr` has a **small sum of 
squares** test (**smallsstest**) that compares the latest evaluated sum of squared
(weighted) residuals to `e4` times the initial sum of squares, where
`e4 <- (100*.Machine$double.eps)^4` is approximately 2.43e-55.

Termination after what may be considered excessive computation is also important.
`nls()` stops after `maxiter` "iterations". The meaning of "iteration" may require
examination of the code. We prefer to use the number of times the residuals or 
the jacobian have been computed and put upper limits on these. Our codes exit (terminate) 
when these limits are reached. Generally we prefer larger limits than the 
default `maxiter = 50` of `nls()` to avoid stopping early.

@@

?? Can we simplify or suggest use of `str()` ??

The output of `nls()` is an object of class "nls" which has the following structure:

>A list of

>`m` &nbsp; &nbsp; &nbsp; &nbsp; &nbsp; &nbsp; &nbsp; &nbsp; &nbsp; &nbsp; &nbsp; &nbsp; &nbsp; &nbsp; &nbsp;&nbsp;  an `nlsModel` object incorporating the model.

>`data`	&nbsp; &nbsp; &nbsp; &nbsp; &nbsp; &nbsp; &nbsp; &nbsp; &nbsp; &nbsp; &nbsp; &nbsp; &nbsp; &nbsp; &nbsp;the expression that was passed to `nls` as the data argument. The actual \newline
$~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~$  data values are present in the environment of the `m` components, e.g., \newline $~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~$ `environment(m$conv)`.

>`call` &nbsp; &nbsp; &nbsp; &nbsp; &nbsp;  &nbsp; &nbsp; &nbsp; &nbsp; &nbsp; &nbsp; &nbsp; &nbsp; the matched call with several components, notably `algorithm`.

>`na.action` &nbsp; &nbsp;  &nbsp; &nbsp; &nbsp; &nbsp; &nbsp; &nbsp; &nbsp; the `"na.action"` attribute (if any) of the model frame.

>`dataClasses` &nbsp; &nbsp; &nbsp; &nbsp; &nbsp; &nbsp; &nbsp; $~$the `"dataClasses"` attribute (if any) of the "`terms`" attribute of the \newline
$~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~$ model frame.

>`model` &nbsp; &nbsp;  &nbsp; &nbsp; &nbsp; &nbsp; &nbsp; &nbsp; &nbsp;  &nbsp; &nbsp; &nbsp; &nbsp;if `model = TRUE`, the model frame.

>`weights` &nbsp; &nbsp; &nbsp; &nbsp; &nbsp; &nbsp; &nbsp; &nbsp; &nbsp; &nbsp; &nbsp; if `weights` is supplied, the weights.

>`convInfo` &nbsp; &nbsp;  &nbsp; &nbsp; &nbsp; &nbsp; &nbsp; &nbsp; &nbsp; &nbsp;&nbsp; a list with convergence information.

>`control` &nbsp; &nbsp; &nbsp; &nbsp; &nbsp; &nbsp; &nbsp; &nbsp; &nbsp; &nbsp; &nbsp; the control `list` used, see the `control` argument.

>`convergence, message` &nbsp;for an `algorithm = "port"` fit only, a convergence code (`0` for  \newline
$~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~$convergence) and message.

> &nbsp; &nbsp; &nbsp; &nbsp; &nbsp; &nbsp; &nbsp; &nbsp; &nbsp; &nbsp; &nbsp; &nbsp; &nbsp; &nbsp; &nbsp; &nbsp; &nbsp; To use these is *deprecated*, as they are available from `convInfo` now.

                  -----------------------------------------------------------

The complexity of this object is a challenge to users. Let us use 
`result` as the returned object from `nls()`, `minpack.lm::nlsLM()` or
`gslnls::gsl_nls()`. For example, the `data` return element is an R symbol. To
actually access the data from this element, we need to use the syntax:

```
eval(parse(text=result$data))
```

However, if the call is made with `model=TRUE`, then there is a returned element
`model` which contains the data, and we can list its contents using:

```
ls(result$model)
```

and if there is an element called `xdata` that can be accessed as
`result$model$xdata`.

By contrast `nlsr::nlxb()` returns

>`coefficients`    A named vector giving the parameter values at the supposed solution.

>`ssquares` $~~~~~$ The sum of squared residuals at this set of parameters.

>`resid` $~~~~~~~~~~$ The residual vector at the returned parameters.

>`jacobian` $~~~~~$ The jacobian matrix (partial derivatives of residuals w.r.t. the 
                parameters) at the \newline
$~~~~~~~~~~~~~~~~~~~~$ returned parameters.

>`feval` $~~~~~~~~~~$ The number of residual evaluations (sum of squares computations) used.

>`jeval` $~~~~~~~~~~$ The number of Jacobian evaluations used.

However, actually looking at the structure of a returned result gives a list of 11
items. The extra 5 are:

```
 $ lower       : num [1:3] -Inf -Inf -Inf
 $ upper       : num [1:3] Inf Inf Inf
 $ maskidx     : int(0) 
 $ weights     : NULL
 $ formula     :Class 'formula'  language y ~ Asym/(1 + exp((xmid - tt)/scal))
  .. ..- attr(*, ".Environment")=<environment: R_GlobalEnv> 
 - attr(*, "class")= chr "nlsr"
```
The result object from `nlsr::nlxb()` is still much smaller than the one `nls()` 
returns. Moreover, `nlxb`
explicitly returns the sum of squares as well as the residual vector and Jacobian. 
The counts of evaluations are also returned.

@@

Using the "gradient" attribute of the output of the Jacobian function to hold
this matrix lets us embed this attribute in the **residual** function as well,
so that the call to `nlfb()` can be made with the same name used for 
both residual and Jacobian function arguments. This programming trick saves a 
lot of trouble for the package developer, but it can be a nuisance for users 
trying to understand the code.

@@

`minpack.lm::nlsLM()` invokes `numericDeriv()` in its local 
version of `nlsModel()`, but it appears to use an internal approximate Jacobian 
code from 
the original Fortran `minpack` code, namely, `lmdif.f`. Such differences in approach
can lead to different behaviour, though in our experience these differences are usually
very minor.

However, minor numerical differences can give annoying results with ill-conditioned
problems.

@@

- Karl Schilling (private communication) provided an example where a model specified
  with the formula `y ~ a * (x ˆ b)` causes `nlsr::nlxb` to fail because the partial
  derivative w.r.t. `b` is `a * (x^b * log(x))`. If there is data for which `x = 0`,
  the derivative is undefined, while the model can be computed. In such cases, 
  we observed that `nls()` and 
  `minpack.lm::nlsLM` found a solution, though this seems to be a lucky accident.

@@

Analytic Jacobian code can be provided to `nls()`, `nlsLM()` and `gsl_nls()`,
and most `selfStart` models that automatically provide starting parameters
also include such code. There is documentation in R
of `selfStart` models, but the construction of them is non-trivial. A number 
of such models are included with base R in `./src/library/stats/R/zzModels.R`,  
with package `nlraa` (@MiguezNLRAA2021) providing a richer set. 
There are also some in the now-archived package `NRAIA`.
These provide the Jacobian in the "gradient" attribute of the "one-sided" formula
that defines
each model, and these Jacobians may be the analytic forms.

The `nls()` function, after
computing the "right hand side" or `rhs` of the residual, then checks to see if the
"gradient" attribute is defined, and, if not, uses `numericDeriv()` to compute a 
Jacobian
into that attribute. This code is within the `nlsModel()` or `nlsModel.plinear()`
functions. The use of analytic Jacobians
almost certainly contributes to the good performance of `nls()` on `selfStart` 
models.

The use of `selfStart` models with
`nlsr` is described in the "Introduction to nlsr" vignette. However, since `nlsr`
generally can use very crude starting values, we have rarely needed them, though
it should be pointed out that our work is primarily diagnostic. If we were carrying
out a large number of similar estimations, then good initial parameters are 
critical to efficiency.

In considering `selfStart` models, we noted that the base-R function `SSlogis`
is intended to solve our example problem **Logistic3T**.  This example is one 
where the `getInitial()` function (used to find
starting values) actually calls `nls()` recursively with a non-default algorithm
to get "starting" parameters. Then it returns to running the default algorithm
unnecessarily. Such circular references are, in our view, prone to creating errors
in code maintenance. 
To provide simpler starting parameters, the function `SSlogisJN`
is now part of package `nlsr`, but is most useful for `nls()`.

Users may also want to use `selfStart` models with explicit starting values 
other than those suggested by `getInitial()`. 
We are also surprised that the 
analytic expressions for the Jacobian ("gradient") in the `SSLogis` R function 
and others save many quantities in "hidden"
variables, i.e., with names preceded by ".". 
These are then not displayed by the `ls()`
command, making them difficult to access by users who may wish to create
their own selfStart model by a copy and edit.
Interactive tools, such as "visual fitting" (@nash1996nonlinear) might be
worth considering as a way to find starting parameters, but we know of no
R capability of this type.

@@

In some cases, we know that parameters cannot be bigger or smaller than some
externally known limits. Cash reserves cannot be negative. Animals have a 
minimum need for water. Airplanes cannot carry more than a known or 
legislated cargo weight. Such limits can be built into models, but there 
are some important details for using the tools in R.

@@

- `nlsLM()` includes bounds in the standard call, but we have observed cases
   where it fails to get the correct answer. From examination of the code,
   we believe the authors have not taken into account all possibilities, though
   it should be noted that all programs have some weakness regarding constrained
   optimization. Software creators have to work with
   assumptions on the extremity of scale that they are willing to countenance,
   and sometimes problems will be outside the scope envisaged.

@@

Bounds on parameters raise some interesting and difficult questions about how 
uncertainty in parameter estimates should be computed or reported. That is, the
traditional "standard errors" are generally taken to imply symmetric intervals about
the point estimate in which the parameter may be expected to be found with some
probability under certain assumptions.
Bounds change those assumptions. Hence, `nlsr::nlxb()` does not compute standard
errors nor their derived statistics when bounds are active.

@@

All four major tools illustrated solve some variant of the Gauss-Newton 
equations. `nls()` uses a modification of an approach suggested by @Hartley1961,
while `nlsr`, `gslnls` and `minpack.lm` use flavours of @Marquardt1963. `gslnls`
offers an accelerated Marquardt method and three alternative methods. We have 
not explored all five methods with various control settings.
Control settings for `nlxb()` or `nlfb()` allow exploration of Hartley and Marquardt 
algorithm variants.

In general, the Levenberg-Marquardt stabilization is important in obtaining 
solutions in methods of the Gauss-Newton family, as `nls()` terminates 
too frequently and unnecessarily with `singular gradient` errors.

@@

`na.action` is an argument to the `nls()` function, but it does not appear 
obviously in the source code, often being handled behind the scenes after 
referencing the option `na.action`. Yet this feature also changes the data 
supplied to our nonlinear least squares solver.

A useful, but possibly dated, description is given in:
*https://stats.idre.ucla.edu/r/faq/how-does-r-handle-missing-values/*.
The typical default action, which can be seen by using the command
`getOption("na.action")`
is `na.omit`. This option removes from computations any observations
containing missing values (i.e. any row of a data frame containing an NA). 
`na.exclude` does much of the same for solver computations, but keeps the rows with NA 
elements so that predictions are in the correct row position. We recommend that 
workers actually test output to verify the behaviour is as wanted.
See *https://stats.stackexchange.com/questions/492955/should-i-use-na-omit-or-na-exclude-in-a-linear-model-in-r*.

As with `subset`, our concern with `na.action` is that users may be unaware of the
effects of an option they may not even be aware has been set. Should `na.fail` be the default?

@@

<!-- ?? Arkajyoti: I think this example may get too long or be too unclear. -->

@@

## Minor complaints

@@

The variable projection method (@Golub1973, @OlearyRust13) is 
generally much more effective than general approaches in finding good solutions 
to nonlinear least squares 
problems when some of the parameters appear linearly. In our logistic examples, 
the asymptote parameters are an illustration.However, setting up the 
calculations, that is, identifying which parameters are linear, is not a trivial
task.

@@

Indexed parameters -- those whose names have subscripts on some element that
is part of the coefficient list -- 
are a useful idea to simplify model specification in some cases. However, they
are also a focus of trouble. Typically we have some common parameters and 
others that are tied to particular cases.

The **man** file for `nls()` includes the following example of a situation in which
parameters are indexed. It also uses the "plinear" option as an added complication.
Here we use a truncated version of the example to save display space.

```{r nlsindx1}
## The muscle dataset in MASS is from an experiment on muscle
## contraction on 21 animals.  The observed variables are Strip
## (identifier of muscle), Conc (Cacl concentration) and Length
## (resulting length of muscle section).
if(! requireNamespace("MASS", quietly = TRUE)) stop("Need MASS pkg")
mm<- MASS::muscle[1:12,] # take only 1st few values of Strip (TRUNCATION OF EXAMPLE)
mm<-droplevels(mm) # remove unused levels after truncation
nlev <- nlevels(mm)
withAutoprint({
  ## The non linear model considered is
  ##       Length = alpha + beta*exp(-Conc/theta) + error
  ## where theta is constant. For now alpha and beta do NOT vary with Strip.
  with(mm, table(Strip)) # 2, 3 or 4 obs per strip
  nl <- nlevels(mm$Strip)
  ## We first use the plinear algorithm to fit an overall model,
  ## ignoring that alpha and beta might vary with Strip.
  musc.1 <- nls(Length ~ cbind(1, exp(-Conc/th)), mm,
                start = list(th = 1), algorithm = "plinear")
  summary(musc.1)

  ## Then we use nls' indexing feature for parameters in non-linear
  ## models to use the conventional algorithm to fit a model in which
  ## alpha and beta vary with Strip.  The starting values are provided
  ## by the previously fitted model.
  ## Note that with indexed parameters, the starting values must be
  ## given in a list (with names):
  b <- coef(musc.1)
  musc.2 <- nls(Length ~ a[Strip] + b[Strip]*exp(-Conc/th), data=mm,
                start = list(a = rep(b[2], nl), b = rep(b[3], nl), th = b[1]))
  summary(musc.2)
})
```

Running the example reveals that the answers for the parameters are NOT indexed. 
<!-- e.g. `coef(musc.2)` is a single level vector of parameters.  -->
That is, we do not see `a[1], a[2], a[3]` but `a1, a2, a3`. 
<!-- This is because the model -->
<!-- must integrate all the parameters because `th` is a common parameter across the index `Strip`. -->
This is no doubt because programming for indexed parameters is extremely challenging.
<!-- However, we believe the current structure could cause confusion and error, and  -->
<!-- propose an alternative approach below. -->

If such models are of enough importance, we believe it would be relatively
straightforward to prepare an external package to provide for their estimation,
including an appropriately formatted output object so the indexing of the 
parameters can be used in other calculations.

@@

@@

Maintainers of packages need suitable tests and use-case examples in order:

- to ensure packages work properly, in particular, giving
  results comparable to or better than the functions they are to replace;
- to test individual solver functions to ensure they work across the range of 
  calling mechanisms, that is, different ways of supplying inputs to the solver(s);
- to pose "silly" inputs to see if these bad inputs are caught by the programs.

Such goals align with the aims of **unit testing** 
(e.g., *https://towardsdatascience.com/unit-testing-in-r-68ab9cc8d211*, @HWtestthat11, @HWdevtools21
and the conventional R package testing tools).

In our work, one of us has developed a working prototype package at *https://github.com/ArkaB-DS/nlsCompare* .
A primary design objective of this is to allow the summarisation of multiple tests in
a compact output. The prototype has a vignette to illustrate its use.

@@

## Should nls() be refactored?

When we started *Improvements to nls()* for the 2021 Google Summer of Code we wanted:

- to provide a packaged version of `nls()` (call it `nlsalt`) coded entirely in R 
  that matches the features in base R or what is packaged in `nlspkg` as described 
  in @PkgFromRbase22. Our intent in providing a purely R version of the code was
  either to replace `nls()` or to serve as a reference version aiding maintenance 
  of C, Fortran, or other components of the software.

- to streamline the overall `nls()` infrastructure. By this we mean a 
  refactoring of the routines so they are better suited to maintenance of both
  the existing `nls()` methods and features as well as new features we or others
  would like to add.

- to explain what we do, either in comments, examples, or separate maintainer 
  documentation. While R and its packages are generally very well documented for
  usage, that documentation
  rarely extends as far as it might for code maintenance. The paucity of such
  documentation is exacerbated by the mixed R/C/Fortran code base,
  where seemingly trivial differences in things like representation of numbers or
  base index for arrays lead to traps for unwary programmers.

These goals echo themes in the recent discussion by @Vidoni21 and accompanying 
commentary.

We found the legacy code and examples prevented us achieving a revised `nls()` 
with all its current capabilities. Unfortunately, `nls()` is part of the base R
distribution, 
so users are drawn to using it even when it is less than ideal. A possible way forward
is to

- use a packaged version of `nls()` to provide legacy features, with notification
  to users that there are other tools available

- if maintainers are willing, to merge and rationalize the capabilities of 
  `minpack.lm`, `nlsr` and `gslnls`

- to encourage the (separate) development of capabilities such as indexed
  parameter models if these are in demand.

This direction does not, of course, resolve the inconsistencies in model 
specification in formulas for nonlinear models, which undoubtedly is an issue
for other modeling tools.

A possible workaround to allow consistency for the `plinear` algorithm 
would be to specify the linear parameters when the `plinear` option is specified. 
For example, we could use `algorithm="plinear(Asym)"` while keeping the general 
model formula from the default specification. This would allow 
for the output of different `algorithm` options to be consistent. However, we 
have not yet tried to code such a change, and welcome discussion from those working
with other packages using model formulas.

A further complication of model specification in R is that some formulas require
the user to surround a term with `I()` to inhibit interpretation of arithmetic 
operators. We would prefer that formulas be usable without this complication.

>>>>

```{r exhobbsfn, eval=TRUE}
hobnlfb<-nlfb(start=stu1, resfn=hobbs.res, jacfn=hobbs.jac)
pshort(hobnlfb) # use print(hobnlfb) for more detail
hobnlm<-nls.lm(par=stu1, fn=hobbs.res, jac=hobbs.jac)
pnlm1(hobnlm)  
hobgsln<-gsl_nls(start=stu1, fn=hobbs.res, y=rep(0,12))
pnlm0(hobgsln)
hobgsl<-gsl_nls(start=stu1, fn=hobbs.res, y=rep(0,12), jac=hobbs.jac)
pnlm0(hobgsl) # using analytic Jacobian
```

@@

The design goals of the different tools may also be revealed in the so-called
"convergence tests" for the iterative solvers. In the manual page for `nls()` 
in R 4.0.0 there was the warning:

@@

with suggested addition of small perturbations to the data. This admits 
`nls()` could not solve well-posed problems unless data is polluted
with errors. Zero-residual problems are not always artificial,
since problems in function approximation and nonlinear equations can be 
approached with nonlinear least squares. Fortunately, a small adjustment
to the "termination test" for the **program**, rather than for 
"convergence" of the underlying **algorithm**, fixes the defect. The test
is the Relative Offset Convergence
Criterion (see @BatesWatts81). This scales an estimated reduction in the loss
function by its current value. If the loss function is very small, we are close
to a zero-divide. Adding a small quantity to the divisor avoids trouble. 
In 2021, one of us (J. Nash) proposed that `nls.control()` have an 
additional parameter `scaleOffset` with a default value of zero. Setting it to a 
small number -- 1.0 is a reasonable choice -- allows small-residual problems 
(i.e., near-exact fits) to be dealt with easily. We call this the
**safeguarded relative offset convergence criterion**, and it has been in `nlsr`
since it was introduced. The default value 
gives the legacy behaviour. This improvement has been in the R distributed code 
since version 4.1.0.

Additional termination tests can be used. `nlsr` has a **small sum of 
squares** test (**smallsstest**) that compares the latest evaluated sum of squared
(weighted) residuals to `e4` times the initial sum of squares, where
`e4 <- (100*.Machine$double.eps)^4` is approximately 2.43e-55.

Termination after what may be considered excessive computation is also important.
`nls()` stops after `maxiter` "iterations". The meaning of "iteration" may require
examination of the code. We terminate execution when the number of residual or 
jacobian evaluations exceed set limits. Generally we prefer larger limits than the 
default `maxiter = 50` of `nls()` to avoid stopping early.

@@

The output of `nls()` is an object of class "nls" which has quite rich structure
described in the manual file or revealed by applying the `str()` function to the
result of `nls()`.

The complexity of this object is a challenge to users. Let us use 
`result` as the returned object from `nls()`, `minpack.lm::nlsLM()` or
`gslnls::gsl_nls()`. For example, the `data` return element is an R symbol. To
actually access the data from this element, we need to use the syntax:
```
eval(parse(text=result$data))
```
However, if the call is made with `model=TRUE`, then there is a returned element
`model` which contains the data, and we can list its contents using:
```
ls(result$model)
```
and if there is an element called `xdata` that can be accessed as
`result$model$xdata`.

By contrast `nlsr::nlxb()` returns a much simpler structure of 11 items
in one-level. Moreover, `nlxb` explicitly returns the sum of squares, the residual vector, Jacobian and counts of evaluations.

@@

Using the "gradient" attribute of the output of the Jacobian function to hold
the Jacobian matrix lets us embed this in the **residual** function as well,
so that the call to `nlfb()` can be made with the same name used for 
both residual and Jacobian function arguments. This programming trick saves a 
lot of trouble for the package developer, but it can be a nuisance for users 
trying to understand the code.

@@

`minpack.lm::nlsLM()` invokes `numericDeriv()` in its local 
version of `nlsModel()`, but it appears to use an internal approximate Jacobian 
code from 
the original Fortran `minpack` code, namely, `lmdif.f`. Such differences in approach
can lead to different behaviour, usually minor, but sometimes annoying with
ill-conditioned problems.

@@

- Karl Schilling (private communication) provided an example where a model specified
  with the formula `y ~ a * (x ˆ b)` causes `nlsr::nlxb` to fail because the partial
  derivative w.r.t. `b` is `a * (x^b * log(x))`. If there is data for which `x = 0`,
  the derivative is undefined, but the model can be computed. In such cases, 
  we observed that `nls()` and 
  `minpack.lm::nlsLM` found a solution, though this seems to be a lucky accident.

@@

Analytic Jacobian code can be provided to all the solvers discussed.
Most `selfStart` models that automatically provide starting parameters
also include such code. There is documentation in R
of `selfStart` models, but their construction is non-trivial. A number 
of such models are included with base R in `./src/library/stats/R/zzModels.R`,  
with package `nlraa` (@MiguezNLRAA2021) providing a richer set. 
There are also some in the now-archived package `NRAIA`.
These provide the Jacobian in the "gradient" attribute of the "one-sided" formula
that defines each model, and these Jacobians are often the analytic forms.

The `nls()` function, after
computing the "right hand side" or `rhs` of the residual, checks to see if the
"gradient" attribute is defined, and, otherwise using `numericDeriv()` to compute a 
Jacobian into that attribute. This code is within the `nlsModel()` or
`nlsModel.plinear()` functions. The use of analytic Jacobians
almost certainly contributes to the good performance of `nls()` on `selfStart` 
models.

The use of `selfStart` models with
`nlsr` is described in the "Introduction to nlsr" vignette. However, since `nlsr`
generally can use very crude starting values, we have rarely needed them, though
it should be pointed out that our work is primarily diagnostic. If we were carrying
out a large number of similar estimations, such initial parameters are 
critical to efficiency.

In considering `selfStart` models, we noted that the base-R function `SSlogis`
is intended to solve problem **Logistic3T** above.  When this function is used
via `getInitial()` to  find
starting values, it actually calls `nls()` with the 'plinear' algorithm 
and finds a solution. It then passes the solution coefficients to the default 
algorithm unnecessarily. Moreover, the double call is, in our view, prone to creating 
errors in code maintenance. 
To provide simpler starting parameters, the function `SSlogisJN`
is now part of package `nlsr`, but is most useful for `nls()`.

Users may also want to profit from the Jacobian code of `selfStart` models 
but supply explicit starting values other than those suggested by `getInitial()`. 
This does not appear to be possible with `nls()`. `nlsr::nlxb()` always requires 
starting parameters, and uses `getInitial()' to find them from the selfStart
model. The same model is used in the formula for the `nlxb()` call.

<!-- tssx <-try(nls(weed ~ SSlogis(tt, Asym, xmid, scale), start=stt1, data=weeddf)) -->
<!-- tssx <-try(nls(weed ~ SSlogis(tt, Asym=1, xmid=1, scale=1),  data=weeddf)) -->
<!-- tssx <-try(nls(weed ~ SSlogis(tt, Asym, xmid, scale),  data=weeddf))summary(tssx) -->

We are also surprised that the 
analytic expressions for the Jacobian ("gradient") in the `SSLogis` R function 
and others save quantities in "hidden"
variables, i.e., with names preceded by ".". 
These are then not displayed by the `ls()`
command, making them difficult to access by users who may wish to create
their own selfStart model by a copy and edit.
Interactive tools, such as "visual fitting" (@nash1996nonlinear) might be
worth considering as another way to find starting parameters, but we know of no
R capability of this type.

@@

In some cases, we know that parameters cannot be bigger or smaller than some
externally known limits. Such limits should be built into models, but there 
are some important details for using the tools in R.

@@

- `nlsLM()` includes bounds in the standard call, but we have observed cases
   where it fails to get the correct answer. From examination of the code,
   we believe the authors have not taken into account all possibilities, though
   all programs have some weakness regarding constrained optimization in that
   programmers have to work with assumptions on the scale of numbers, 
   and some problems will be outside the scope envisaged.

@@

Bounds on parameters raise some interesting questions about how 
uncertainty in parameter estimates should be computed or reported. That is, the
traditional "standard errors" are generally taken to imply symmetric intervals about
the point estimate in which the parameter may be expected to be found with some
probability under certain assumptions.
Bounds change those assumptions. Hence, `nlsr::nlxb()` does not compute standard
errors nor their derived statistics when bounds are active.

@@

All four major tools illustrated solve some variant of the Gauss-Newton 
equations. `nls()` uses a modification of an approach suggested by @Hartley1961,
while `nlsr`, `gslnls` and `minpack.lm` use flavours of @Marquardt1963. `gslnls`
offers an accelerated Marquardt method and three alternative methods which we
have not investigated.
Control settings for `nlxb()` or `nlfb()` allow exploration of Hartley and Marquardt 
algorithm variants. In general, the Levenberg-Marquardt stabilization 
is important in obtaining 
solutions in methods of the Gauss-Newton family, as `nls()` terminates 
too frequently and unnecessarily with `singular gradient` errors.

@@

`na.action` is an argument to the `nls()` function, but it does not appear 
obviously in the source code, often being handled behind the scenes after 
referencing the option `na.action`, yet this feature also changes the data 
supplied to our nonlinear least squares solver. 
A useful, but possibly dated, description is given in:
*https://stats.idre.ucla.edu/r/faq/how-does-r-handle-missing-values/*.
The typical default action, which can be seen by using the command
`getOption("na.action")`
is `na.omit`. This option removes from computations any observations
containing missing values (i.e. any row of a data frame containing an NA). 
`na.exclude` does much of the same for solver computations, but keeps the rows with 
NA elements so that predictions are in the correct row position. We recommend that 
workers actually test output to verify the behaviour is as wanted.
See *https://stats.stackexchange.com/questions/492955/should-i-use-na-omit-or-na-exclude-in-a-linear-model-in-r*.
As with `subset`, our concern with `na.action` is that users may be unaware of the
effects of an option they may not even be aware has been set. Should `na.fail` be the default?

@@

@@

## Minor issues with nonlinear least-squares tools

@@

The variable projection method (@Golub1973, @OlearyRust13) is 
generally much more effective than general approaches in finding good solutions 
to nonlinear least squares 
problems when some of the parameters appear linearly. In our logistic examples, 
the asymptote parameters are an illustration. However, identifying which parameters
are linear and communicating this information to estimating functions 
is not a trivial task.

@@

Some models have several common parameters and 
others that are tied to particular cases. 
The **man** file for `nls()` includes an example of a situation in which
parameters are indexed but which uses the "plinear" option as an added complication.
Running this example reveals that the answers for the parameters are NOT indexed
as in a vector. 
<!-- e.g. `coef(musc.2)` is a single level vector of parameters.  -->
That is, we do not see `a[1], a[2], a[3]` but `a1, a2, a3`. 
<!-- This is because the model -->
<!-- must integrate all the parameters because `th` is a common parameter across the index `Strip`. -->
This is no doubt because programming for indexed parameters is challenging.
<!-- However, we believe the current structure could cause confusion and error, and  -->
<!-- propose an alternative approach below. -->
We note that there are capabilities in packages for mixed-effects modelling
such as `nlme` (@p-nlme), `bbmle` (@p-bbmle), and `lme4` (p-lme4) for 
estimating models of such type.

@@

<!-- ===```{r nlsindx1} -->
<!-- ## The muscle dataset in MASS is from an experiment on muscle -->
<!-- ## contraction on 21 animals.  The observed variables are Strip -->
<!-- ## (identifier of muscle), Conc (Cacl concentration) and Length -->
<!-- ## (resulting length of muscle section). -->
<!-- if(! requireNamespace("MASS", quietly = TRUE)) stop("Need MASS pkg") -->
<!-- mm<- MASS::muscle[1:12,] # take only 1st few values of Strip (TRUNCATION OF EXAMPLE) -->
<!-- mm<-droplevels(mm) # remove unused levels after truncation -->
<!-- nlev <- nlevels(mm) -->
<!-- withAutoprint({ -->
<!--   ## The non linear model considered is -->
<!--   ##       Length = alpha + beta*exp(-Conc/theta) + error -->
<!--   ## where theta is constant. For now alpha and beta do NOT vary with Strip. -->
<!--   with(mm, table(Strip)) # 2, 3 or 4 obs per strip -->
<!--   nl <- nlevels(mm$Strip) -->
<!--   ## We first use the plinear algorithm to fit an overall model, -->
<!--   ## ignoring that alpha and beta might vary with Strip. -->
<!--   musc.1 <- nls(Length ~ cbind(1, exp(-Conc/th)), mm, -->
<!--                 start = list(th = 1), algorithm = "plinear") -->
<!--   summary(musc.1) -->

<!--   ## Then we use nls' indexing feature for parameters in non-linear -->
<!--   ## models to use the conventional algorithm to fit a model in which -->
<!--   ## alpha and beta vary with Strip.  The starting values are provided -->
<!--   ## by the previously fitted model. -->
<!--   ## Note that with indexed parameters, the starting values must be -->
<!--   ## given in a list (with names): -->
<!--   b <- coef(musc.1) -->
<!--   musc.2 <- nls(Length ~ a[Strip] + b[Strip]*exp(-Conc/th), data=mm, -->
<!--                 start = list(a = rep(b[2], nl), b = rep(b[3], nl), th = b[1])) -->
<!--   summary(musc.2) -->
<!-- }) -->
<!-- === ``` -->

@@

Maintainers of packages need suitable tests and use-case examples in order
- to ensure packages work properly, in particular, giving
results comparable to or better than the functions they are to replace. 
- to test individual solver functions to ensure they work across the range of 
  calling mechanisms, that is, different ways of supplying inputs to the solver(s);
- to pose "silly" inputs to see if these bad inputs are caught by the programs.

Such goals align with the aims of **unit testing** 
(e.g., *https://towardsdatascience.com/unit-testing-in-r-68ab9cc8d211*, @HWtestthat11, @HWdevtools21
and the conventional R package testing tools).
In our work, one of us has developed a working prototype package at *https://github.com/ArkaB-DS/nlsCompare* .
A primary design objective of this is to allow the summarisation of multiple tests in
a compact output. The prototype has a vignette to illustrate its use.

@@

<!-- ## Should nls() be refactored? -->

<!-- When we started *Improvements to nls()* for the 2021 Google Summer of Code we wanted: -->

<!-- - to provide a packaged version of `nls()` (call it `nlsalt`) coded entirely in R  -->
<!--   that matches the features in base R or what is packaged in `nlspkg` as described  -->
<!--   in @PkgFromRbase22. Our intent in providing a purely R version of the code was -->
<!--   either to replace `nls()` or to serve as a reference version aiding maintenance  -->
<!--   of C, Fortran, or other components of the software. -->

<!-- - to streamline the overall `nls()` infrastructure. By this we mean a  -->
<!--   refactoring of the routines so they are better suited to maintenance of both -->
<!--   the existing `nls()` methods and features as well as new features we or others -->
<!--   would like to add. -->

<!-- - to explain what we do, either in comments, examples, or separate maintainer  -->
<!--   documentation. While R and its packages are generally very well documented for -->
<!--   usage, that documentation -->
<!--   rarely extends as far as it might for code maintenance. The paucity of such -->
<!--   documentation is exacerbated by the mixed R/C/Fortran code base, -->
<!--   where seemingly trivial differences in things like representation of numbers or -->
<!--   base index for arrays lead to traps for unwary programmers. -->

<!-- These goals echo themes in the recent discussion by @Vidoni21 and accompanying  -->
<!-- commentary. -->

<!-- We found the legacy code and examples prevented us achieving a revised `nls()`  -->
<!-- with all its current capabilities. Unfortunately, `nls()` is part of the base R -->
<!-- distribution,  -->
<!-- so users are drawn to using it even when it is less than ideal. A possible way forward -->
<!-- is to  -->

<!-- - use a packaged version of `nls()` to provide legacy features, with notification -->
<!--   to users that there are other tools available -->

<!-- - if maintainers are willing, to merge and rationalize the capabilities of  -->
<!--   `minpack.lm`, `nlsr` and `gslnls` -->

<!-- - to encourage the (separate) development of capabilities such as indexed -->
<!--   parameter models if these are in demand. -->

<!-- This direction does not, of course, resolve the inconsistencies in model  -->
<!-- specification in formulas for nonlinear models, which undoubtedly is an issue -->
<!-- for other modeling tools.   -->
